---
title: "Particulate Matter Exploratory Data Analysis"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The overall goal of this project is to explore the National Emissions Inventory database and see what it say about fine particulate matter pollution in the United states over the 10-year period 1999–2008. 

### Steps

Pick libraries to use in the project

```{r message=TRUE, warning=TRUE}
library(ggplot2)
library(dplyr)
library(readr)
library(lubridate)
```

Download and extract data set from EPA National Emissions Inventory (USA)

```{r}
download.file(url = "https://d396qusza40orc.cloudfront.net/exdata%2Fdata%2FNEI_data.zip", destfile = "mydata.zip")
unzip("mydata.zip", exdir = "./data")
file.remove("mydata.zip")
```

Read tables into the Environment

```{r}
# National Emissions Inventory (NEI)
nei <- read_rds("./data/summarySCC_PM25.rds")
# Source Classification Code (SCC)
scc <- read_rds("./data/Source_Classification_Code.rds")
```

See a glimpse of the content

```{r}
glimpse(nei)
```

```{r}
glimpse(scc)
```

### Questions:
*Have total emissions from PM2.5 decreased in the United States from 1999 to 2008? Using the base plotting system, make a plot showing the total PM2.5 emission from all sources for each of the years 1999, 2002, 2005, and 2008.*

Look for years tracked in inventory

```{r}
unique(nei$year)
```

Filter total emissions for each year

```{r}
total_emissions <- nei %>% 
    group_by(year) %>% 
    summarize(
        sum_emissions = sum(Emissions)
    )

total_emissions
```

Change column *sum_emissions* to integer

```{r}
total_emissions <- total_emissions %>% 
    mutate_at(vars(sum_emissions), as.integer)
total_emissions
```

Make a barplot with the sum of emissions by each year

```{r echo=FALSE, message=FALSE, warning=FALSE}
par(mfrow = c(1, 1))
plot(
    x = total_emissions$year, 
    y = total_emissions$sum_emissions, 
    type = "l", 
    lwd = 2, 
    col = "blue", 
    ylim = c(0, 8e06),
    xlab = "Year",
    ylab = expression("PM"[2.5]),
    main = "Particulate Matter Emissions"
)
abline(
    lm(
        data = total_emissions, 
        formula = sum_emissions ~ year
    ),
    lty = 2
)

dev.copy(png, "./images/plot1.png")
dev.off()
```

-----

*Have total emissions from PM2.5 decreased in the Baltimore City, Maryland (\color{red}{\verb|fips == "24510"|}fips=="24510") from 1999 to 2008?*

Filter sum of emissions in **Baltimore City**

```{r}
nei_baltimore <- nei %>% 
    filter(fips == "24510") %>% 
    group_by(year) %>% 
    summarize(
        sum_emissions = sum(Emissions)
    )

nei_baltimore
```

Set the canvas to 1 plot and plot the results with a tendency line

```{r echo=FALSE, message=FALSE, warning=FALSE}
par(mfrow = c(1, 1))
plot(
    x = nei_baltimore$year, 
    y = nei_baltimore$sum_emissions, 
    type = "l", 
    lwd = 2, 
    col = "green", 
    ylim = c(0, 3500),
    xlab = "Year",
    ylab = expression("PM"[2.5]),
    main = "Particulate Matter Emissions in Baltimore"
)
abline(
    lm(
        data = baltimore, 
        formula = sum_emissions ~ year
    ),
    lty = 2
)

dev.copy(png, "./images/plot2.png")
dev.off()
```

-----

*Of the four types of sources indicated by the \color{red}{\verb|type|}type (point, nonpoint, onroad, nonroad) variable, which of these four sources have seen decreases in emissions from 1999–2008 for Baltimore City? Which have seen increases in emissions from 1999–2008?*

Change NEI data set to work as a tibble (compatible with the *tidyverse*) and filter emissions from **Baltimore** by each type

```{r}
nei <- as_tibble(nei)

pm_type_baltimore <- nei %>% 
    filter(fips == "24510") %>% 
    group_by(type, year) %>%
    summarize(
        emissions = sum(Emissions)
    ) %>% 
    arrange(year, type)
```

Plot the results

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(pm_type_baltimore, aes(year, emissions, group = type, color = type)) +
    geom_point(na.rm = TRUE) +
    geom_smooth() +
    labs(
        x = "Year",
        y = expression("PM"[2.5]),
        color = "Type",
        title = "Emissions in Baltimore City",
        subtitle = "(1999-2008)",
        shape = ""
    ) +
    geom_smooth(
        method = "lm", 
        lty = 2, 
        se = FALSE,
        lwd = 0.2
    ) +
    theme_minimal()

dev.copy(png, "./images/plot3.png")
dev.off()
```

-----

*Across the United States, how have emissions from coal combustion-related sources changed from 1999–2008?*



```{r}
tscc <- as.tibble(scc)

scc_coal <- tscc %>% 
    filter(grepl("coal", Short.Name, ignore.case = TRUE))
total_coal <- nei %>% 
    filter(SCC %in% scc_coal$SCC) %>% 
    group_by(year, type) %>% 
    summarize(
        emissions = sum(Emissions)
    )

print(total_coal)
```

Plot the results

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(total_coal, aes(year, emissions, group = type, color = type)) +
    geom_point(na.rm = TRUE) +
    geom_smooth(
        method = "lm", 
        lty = 2, 
        se = FALSE,
        color = "grey",
        lwd = .5
    ) +
    geom_smooth() +
    labs(
        x = "Year",
        y = "CCE (ton)",
        color = "Type",
        title = "Total Coal Emissions",
        subtitle = "(1999-2008)"
    ) +
    theme_minimal()
    

dev.copy(png, "./images/plot4.png")
dev.off()
```

