---
title: "Particulate Matter Exploratory Data Analysis"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

> The overall goal of this project is to explore the National Emissions Inventory database and see what it say about fine particulate matter pollution in the United states over the 10-year period 1999–2008. 

## Tools Used

- R language compiler
- R base graphic devices
- Tidyverse library packages
- RMarkdown library package

## Files

  - **Raw data set**: [Environmental Protection Agency National Emissions Inventory](https://d396qusza40orc.cloudfront.net/exdata%2Fdata%2FNEI_data.zip)
  - **[CODEBOOK.md](https://github.com/vcwild/pm-emissions/blob/master/CODEBOOK.md)**: step-by-step book explaining the code processing.
  - **[plot<sub>x</sub>.png](https://github.com/vcwild/pm-emissions/tree/master/images)**: the plotted files
  - **[plot<sub>n</sub>.R](https://github.com/vcwild/pm-emissions/tree/master/scripts)**: the scripts that produces the [plotted images](https://github.com/vcwild/pm-emissions/tree/master/images)
  - **script.Rmd**: the script to build the project from source

## Build from Source

Execute the following command in the terminal

```{r eval=FALSE, include=FALSE}
R
library(rmarkdown)
rmarkdown::render("script.Rmd", "html_document")
```

## CODEBOOK

### Steps

Pick libraries to use in the project

```{r message=TRUE, warning=TRUE}
library(tidyverse)
```

Download and extract data set from EPA National Emissions Inventory (USA)

```{r, cache=TRUE}
download.file(url = "https://d396qusza40orc.cloudfront.net/exdata%2Fdata%2FNEI_data.zip", destfile = "mydata.zip")
unzip("mydata.zip", exdir = "./data")
file.remove("mydata.zip")
```

Read tables into the Environment

```{r, cache=TRUE}
# National Emissions Inventory (NEI)
nei <- read_rds("./data/summarySCC_PM25.rds")
# Source Classification Code (SCC)
scc <- read_rds("./data/Source_Classification_Code.rds")
```

See a glimpse of the content

```{r}
glimpse(nei)
```

```{r}
glimpse(scc)
```

### Questions:
*Have total emissions from PM2.5 decreased in the United States from 1999 to 2008? Using the base plotting system, make a plot showing the total PM2.5 emission from all sources for each of the years 1999, 2002, 2005, and 2008.*

-----

Look for years tracked in inventory

```{r}
unique(nei$year)
```

Filter total emissions for each year

```{r}
total_emissions <- nei %>% 
    group_by(year) %>% 
    summarize(
        sum_emissions = sum(Emissions)
    )

total_emissions
```

Change column *sum_emissions* to integer

```{r}
total_emissions <- total_emissions %>% 
    mutate_at(vars(sum_emissions), as.integer)
total_emissions
```

Make a barplot with the sum of emissions by each year

```{r plot1, echo=TRUE, message=FALSE, warning=FALSE}
par(mfrow = c(1, 1))
plot(
    x = total_emissions$year, 
    y = total_emissions$sum_emissions, 
    type = "l", 
    lwd = 2, 
    col = "blue", 
    ylim = c(0, 8e06),
    xlab = "Year",
    ylab = expression("Total PM"[2.5]*" emission (ton)"),
    main = "Particulate Matter Emissions"
)
abline(
    lm(
        data = total_emissions, 
        formula = sum_emissions ~ year
    ),
    lty = 2
)

dev.copy(png, "./images/plot1.png")
dev.off()
```

-----

*Have total emissions from PM2.5 decreased in the Baltimore City, Maryland (\color{red}{\verb|fips == "24510"|}fips=="24510") from 1999 to 2008?*

-----

Filter sum of emissions in **Baltimore City**

```{r}
nei_baltimore <- nei %>% 
    filter(fips == "24510") %>% 
    group_by(year) %>% 
    summarize(
        sum_emissions = sum(Emissions)
    )

nei_baltimore
```

Set the canvas to 1 plot and plot the results with a tendency line

```{r plot2, echo=TRUE, message=FALSE, warning=FALSE}
par(mfrow = c(1, 1))
plot(
    x = nei_baltimore$year, 
    y = nei_baltimore$sum_emissions, 
    type = "l", 
    lwd = 2, 
    col = "green", 
    ylim = c(0, 3500),
    xlab = "Year",
    ylab = expression("Total PM"[2.5]*" emission (ton)"),
    main = "Particulate Matter Emissions in Baltimore"
)
abline(
    lm(
        data = nei_baltimore, 
        formula = sum_emissions ~ year
    ),
    lty = 2
)

dev.copy(png, "./images/plot2.png")
dev.off()
```

-----

*Of the four types of sources indicated by the \color{red}{\verb|type|}type (point, nonpoint, onroad, nonroad) variable, which of these four sources have seen decreases in emissions from 1999–2008 for Baltimore City? Which have seen increases in emissions from 1999–2008?*

-----

Change NEI data set to work as a tibble (compatible with the *tidyverse*) and filter emissions from **Baltimore** by each type

```{r}
nei <- as_tibble(nei)

pm_type_baltimore <- nei %>% 
    filter(fips == "24510") %>% 
    group_by(type, year) %>%
    summarize(
        emissions = sum(Emissions)
    ) %>% 
    arrange(year, type)
```

Plot the results

```{r plot3, echo=TRUE, message=FALSE, warning=FALSE}
ggplot(pm_type_baltimore, aes(year, emissions, group = type, color = type)) +
    geom_point(na.rm = TRUE) +
    geom_smooth() +
    labs(
        x = "Year",
        y = expression("Total PM"[2.5]*" emission (ton)"),
        color = "Type",
        title = "Emissions in Baltimore City",
        subtitle = "(1999-2008)",
        shape = ""
    ) +
    geom_smooth(
        method = "lm", 
        lty = 2, 
        se = FALSE,
        lwd = 0.2
    ) +
    theme_minimal()

dev.copy(png, "./images/plot3.png")
dev.off()
```

-----

*Across the United States, how have emissions from coal combustion-related sources changed from 1999–2008?*

-----

Set SCC as tibble. 

Filter coal occurrences in SCC and use it as parameter to filter coal occurrences in NEI.

From filtered NEI get the sum of coal emissions by year and type.

```{r}
tscc <- as.tibble(scc)

scc_coal <- tscc %>% 
    filter(grepl("coal", Short.Name, ignore.case = TRUE))

total_coal <- nei %>% 
    filter(SCC %in% scc_coal$SCC) %>% 
    group_by(year, type) %>% 
    summarize(
        emissions = sum(Emissions)
    )

print(total_coal)
```

Plot the results in scatterplot and reggression line models

```{r plot4, echo=TRUE, message=FALSE, warning=FALSE}
ggplot(total_coal, aes(year, emissions, group = type, color = type)) +
    geom_point(na.rm = TRUE) +
    geom_smooth(
        method = "lm", 
        lty = 2, 
        se = FALSE,
        color = "grey",
        lwd = .5
    ) +
    geom_smooth() +
    labs(
        x = "Year",
        y = "CCE (ton)",
        color = "Type",
        title = "Total Coal Emissions",
        subtitle = "(1999-2008)"
    ) +
    theme_minimal()

dev.copy(png, "./images/plot4.png")
dev.off()
```

-----

*How have emissions from motor vehicle sources changed from 1999–2008 in Baltimore City?*

-----

```{r}
scc_vehicle <- tscc %>% 
    filter(grepl("vehicle", SCC.Level.Two, ignore.case = TRUE))

total_vehicles_baltimore <- nei %>% 
    filter(SCC %in% scc_vehicle$SCC, fips == "24510") %>% 
    group_by(year, type) %>% 
    summarize(
        emissions = sum(Emissions)
    )

total_vehicles_baltimore
```

```{r plot5, echo=TRUE, message=FALSE, warning=FALSE}
ggplot(total_vehicles_baltimore, aes(year, emissions, group = type, color = type)) +
    geom_point(na.rm = TRUE) +
    geom_smooth(
        method = "lm", 
        lty = 2, 
        se = FALSE,
        color = "grey",
        lwd = .5
    ) +
    geom_smooth() +
    labs(
        x = "Year",
        y = expression("Total PM"[2.5]*" emission (ton)"),
        color = "Type",
        title = "Total Vehicle Emissions in Baltimore City",
        subtitle = "(1999-2008)"
    ) +
    theme_minimal()

dev.copy(png, "./images/plot5.png")
dev.off()
```

-----

*Compare emissions from motor vehicle sources in Baltimore City with emissions from motor vehicle sources in Los Angeles County, California (\color{red}{\verb|fips == "06037"|}fips=="06037"). Which city has seen greater changes over time in motor vehicle emissions?*

-----

Set the total Vehicle emissions in NEI by using the SCC vehicles vector

```{r}
total_vehicles_ba_la <- nei %>% 
    filter(SCC %in% scc_vehicle$SCC, fips == c("24510", "06037")) %>% 
    group_by(fips, year, type) %>% 
    summarize(
        emissions = sum(Emissions)
    ) %>% 
    mutate(
        city = if_else(fips == "24510", "Baltimore", "Los Angeles")
    )

glimpse(total_vehicles_ba_la)
```

Plot disaggregate

```{r plot6_1, echo=TRUE, message=FALSE, warning=FALSE}
ggplot(total_vehicles_ba_la, aes(year, emissions, group = type, color = type)) +
    geom_point(na.rm = TRUE) +
    geom_line(aes(year, emissions)) +
    facet_grid(city ~ .) +
    geom_smooth(
        method = "lm", 
        lty = 2, 
        se = FALSE,
        color = "grey",
        lwd = .5
    ) +
    labs(
        x = "Year",
        y = expression("Total PM"[2.5]*" emission (ton)"),
        color = "Type",
        title = "Disaggregate Vehicle Emissions Baltimore x Los Angeles",
        subtitle = "(1999-2008)"
    )

dev.copy(png, "./images/plot6_1.png")
dev.off()
```

Using the same method, aggregate the sum of emissions of emissions by each county

```{r}
sum_vehicles_ba_la <- nei %>% 
    filter(SCC %in% scc_vehicle$SCC, fips == c("24510", "06037")) %>% 
    group_by(fips, year) %>% 
    summarize(
        emissions = sum(Emissions)
    ) %>% 
    mutate(
        city = if_else(fips == "24510", "Baltimore", "Los Angeles")
    ) %>% 
    arrange(year)

sum_vehicles_ba_la
```

Plot the total vehicle emissions comparison

```{r plot6_2, echo=TRUE, message=FALSE, warning=FALSE}
ggplot(sum_vehicles_ba_la, aes(year, emissions, group = city, color = city)) +
    geom_point(na.rm = TRUE) +
    geom_smooth(aes(year, emissions)) +
    geom_smooth(
        method = "lm", 
        lty = 2, 
        se = FALSE,
        color = "grey",
        lwd = .5
    ) +
    labs(
        x = "Year",
        y = expression("Total PM"[2.5]*" emission (ton)"),
        color = "County",
        title = "Total Vehicle Emissions Baltimore x Los Angeles",
        subtitle = "(1999-2008)"
    ) +
    theme_minimal()

dev.copy(png, "./images/plot6_2.png")
dev.off()
```

